Nearest Neighbour Methods for Length of Stay Prediction

### 2015-02-02 papers/paper.tex
Paper: slowly coming along, paragraph by paragraph

\section{Introduction}
Hospitals often have limited funding, and therefore managing and utilising
their resources efficiently has always been a priority \cite{Walczak2003}.
One key measurement of hospital activity, health care utilisation, and hospital
cost is the patient length of stay \citep{Omachonu2004,Ng2006}. The length of
stay has been used in a wide variety of medical domains, such as burns
 \citep{Yang2010}, intensive
care \citep{Tu1993,Harper2005,Perez2006,Dybowski1996},
psychiatry \citep{Lowell1997}, and acute pancreatitis \citep{Pofahl1998}, in
order to assist hospitals in allocating beds and other patient management
resources as efficiently as possible. However, given the many different
conditions that hospitals manage, the length of stay of a patient is
not easy to predict \citep{Walczak2003}, especially upon a patient's admission.
Many methods have been explored in the literature, from the realm of statistics
to techniques in data mining, and they have been explored for many different
medical domains.

One of the more well-understood and widely used statistical methods for
predicting the length of stay is \textit{logistic regression} \citep{Tu1996}.
Logistic regression models are straightforward to construct and use via the
help of statistical software packages such as SAS, and
also make the contribution of each independent variable to the final
prediction explicit. Thus they is easy to interpret when compared to more
``black box'' approaches such as artificial neural networks \citep{Adams2012}.
A wide variety of other statistical approaches have been proposed, notably:
various scoring systems from univariate analysis \citep{Adams2012,Lavoie2005},
linear regression \citep{Yang2010}, exponential models \citep{Clark2007},
survival analysis \citep{Vasilakis2005}, and Markov
models \citep{Perez2006,Jain1989,Kapadia2000}.

Out of the development of the various statistically-motivated models, data
mining techniques, particularly artificial neural networks,
began to become more widely adopted following a number of
early applications to length of stay prediction in intensive
care \citep{Tu1993,Dybowski1996}, in addition to disease diagnosis and mortality
prediction \citep{Silva2006}. Artificial neural networks are able to detect
complex and non-linear relationships between input and output variables and can
be developed using many different techniques. Much work has been done in using
neural networks to predict patient length of stay in various medical conditions
such as acute pancreatitis \citep{Walczak2003}, gastroenteritis \citep{Ng2006},
surgical intensive care \citep{Buchman1994,Tu1993} and
psychiatry \citep{Lowell1997}. Other data mining techniques have also been
applied, such as support vector machines and tree-based
methods \citep{Harper2005}.

Given the benefits to the hospital of accurate length of stay prediction,
this research aims to build upon improving length of stay prediction for trauma
patients by utilising state-of-the-art data mining techniques. Being able to
predict the length of stay for trauma patients is especially crucial because of
the scarcity of patient care resource in this area \citep{Yang2010}.
Additionally,
despite the wealth of research into length of stay prediction in various
medical domains, there has been little work done in generalising models from
one domain to another, or to predict a combination of outcomes (not only
length of stay, but mortality, or rehabilitation requirement). Being able to
do this would mean that hospital staff are better equipped to handle patient
cases and prepare for possible critical outcomes well in advance, resulting
in time and resources saved for both hospital and patient.

>>>>

\section{Introduction}
Hospitals often have limited funding, and therefore managing and utilising
their resources efficiently has always been a priority \cite{Walczak2003}.
One key measurement of hospital activity, health care utilisation and hospital
cost is the patient length of stay \citep{Omachonu2004,Ng2006}.
The length of
stay has been used in a wide variety of medical domains, such as burns
 \citep{Yang2010}, intensive
care \citep{Tu1993,Harper2005,Perez2006,Dybowski1996},
psychiatry \citep{Lowell1997}, and acute pancreatitis \citep{Pofahl1998}, in
order to assist hospitals in allocating beds and other patient management
resources as efficiently as possible. However, given the many different
conditions that hospitals manage, the length of stay of a patient is
not easy to predict \citep{Walczak2003}, especially upon a patient's admission.
Many methods have been explored in the literature, from the realm of statistics
to techniques in data mining, and they have been explored for many different
medical domains.

One of the more well-understood and widely used statistical methods for
predicting the length of stay is \textit{logistic regression} \citep{Tu1996}.
Logistic regression models are straightforward to construct and use via the
help of statistical software packages such as
SAS\footnote{SAS Institute, Cary, NC, USA}, and
also make the contribution of each independent variable to the final
prediction explicit. Thus they are easier to interpret when compared to more
``black box'' approaches such as artificial neural networks \citep{Adams2012}.
A wide variety of other statistical approaches have been proposed, notably:
various scoring systems from univariate analysis \citep{Adams2012,Lavoie2005},
linear regression \citep{Yang2010}, exponential models \citep{Clark2007},
survival analysis \citep{Vasilakis2005}, and Markov
models \citep{Perez2006,Jain1989,Kapadia2000}.

Out of the development of the various statistically-motivated models, data
mining techniques, particularly artificial neural networks,
began to be more widely adopted following a number of
early applications to length of stay prediction in intensive
care \citep{Tu1993,Dybowski1996}, in addition to disease diagnosis and mortality
prediction \citep{Silva2006}. Artificial neural networks are able to detect
complex and non-linear relationships between input and output variables and can
be developed using many different techniques. Much work has been done in using
neural networks to predict patient length of stay in various medical conditions
such as acute pancreatitis \citep{Walczak2003}, gastroenteritis \citep{Ng2006},
surgical intensive care \citep{Buchman1994,Tu1993} and
psychiatry \citep{Lowell1997}. Other data mining techniques have also been
applied, such as support vector machines and tree-based
methods \citep{Harper2005}.

One of the major drawbacks of sophisticated data mining techniques is the high
level of difficulty required in interpreting their predictions and the factors
affecting those predictions. This is
particularly true of artificial neural networks and the state-of-the-art
support vector machine, which are able to achieve very high accuracy but can be
non-intuitive to understand. Medical domain experts who use the predictions of
trained learning algorithms need to understand how the algorithm reached its
decision before they are able to confidently use it to aid their clinical
decision-making.

In this paper we present the application of nearest neighbour methods
for predicting the length of stay category for patients admitted to a trauma
ward. The categories are \textit{less than or equal to 2 days} or
\textit{greater than 2 days}, which have been used in previous work on length
of stay prediction in the same medical domain \citep{Dinh2013a}.
Nearest neighour methods have not been studied in predicting the length of stay,
and we believe they are much more intuitive to
understand than techniques such as support vector machines.
Additionally, they can be adjusted to suit particular medical domains by
changing their similarity function. We compare the performance of the basic
nearest neighbour algorithm, as well as an extension (K*), against the
performance
of previous work which used logistic regression. We also propose a novel
extension to the standard nearest neighbour algorithm, Ranked Distance, and
test its performance against the other classifiers. In addition, we apply a
range of feature selection methods, as they have not been
thoroughly studied in existing work. Finally, we compare the performance of
all of these with the state-of-the-art support vector machines and the
commonly-used artificial neural networks.

We found that the standard 1-nearest neighbour classifier achieved the highest
accuracy of 77.81\%, an improvement of 2.75\% over previous work using the same
data set carried out by Dinh et al.\citep{Dinh2013a}.
